---
title: "JSC370 Final Project"
author: "Teresa Zhao"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, warning=FALSE, message=FALSE)
```

# {.tabset}

## Project Description 

### Introduction

Urban traffic congestion is a key urban planning challenge. This project investigates which factors influence traffic volume in Toronto, using data from 2018–2022. While my midterm focused on fuel prices alone, this final project expands the analysis to explore additional variables such as time of day, weekday/weekend patterns, and the COVID-19 pandemic period.

### Research Question

What are the key factors that influence traffic volume in Toronto?

### Data Sources

Traffic Volume: Toronto Open Data (15-minute intervals)

Fuel Prices: Open Canada (monthly taxed fuel prices in Toronto)

## Interactive Plots

```{r eval=FALSE}
install.packages("data.table")
install.packages("patchwork")
install.packages("gridExtra")
install.packages("forcats")
```

```{r, echo=FALSE, eval=TRUE}
library(httr)
library(jsonlite)
library(dplyr)
library(data.table)
library(patchwork)
library(gridExtra)
library(ggplot2)
library(forcats)
library(rpart)
library(rpart.plot)
library(randomForest)
library(gbm)
library(xgboost)
library(caret)
library(doParallel)
library(plotly)
library(leaflet)
library(knitr)
library(mgcv)
```

```{r}
# Define API endpoint (replace {resource_id} with actual resource ID)
resource_id <- "19499837-c088-448f-8c80-f210a75aecb4"
api_url <- paste0("https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/traffic-volumes-midblock-vehicle-speed-volume-and-classification-counts/resource/", resource_id, "/download/svc_raw_data_volume_2015_2019.csv")

# Send GET request
response <- GET(api_url)
dat_2018 <- fread(httr::content(response, as = "text"))
```

```{r}
# Define the correct resource ID for the 2019-2024 volume dataset
resource_id_2024 <- "8581f033-82ee-49fa-b18e-d112d92a6e7e"

# Construct the API request URL
api_url_2024 <- paste0("https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/traffic-volumes-midblock-vehicle-speed-volume-and-classification-counts/resource/", resource_id_2024, "/download/svc_raw_data_volume_2019_2024.csv")

# Send the GET request
response_2024 <- GET(api_url_2024)

# Read the CSV data into R
dat_traffic_2024 <- fread(httr::content(response_2024, as = "text"))
```

```{r}
api_url <- "https://ontario.ca/v1/files/fuel-prices/canadianpumppricesall.csv"
fuel_prices <- fread(api_url)
```

```{r}
dat_filtered <- dat_2018[time_start >= "2018-01-01 00:00:00"]
```

```{r}
# Ensure both datasets have the same column names before merging
dat_traffic_combined <- rbind(dat_filtered, dat_traffic_2024, fill = TRUE)
```

```{r}
# Extract year and month from time_start in dat_traffic_combined
dat_traffic_combined <- dat_traffic_combined %>%
  mutate(month = format(as.Date(time_start), "%Y-%m"))

# Extract year and month from Date in fuel_prices
fuel_prices <- fuel_prices %>%
  mutate(month = format(as.Date(Date), "%Y-%m"))

# Merge datasets based on month
merged_data <- merge(dat_traffic_combined, fuel_prices, by = "month", all = FALSE, allow.cartesian = TRUE)
```

```{r}
merged_data <- merged_data %>%
  mutate(actual_date = as.Date(time_start))
```

```{r}
merged_data <- merged_data %>%
  mutate(
    location_name = as.factor(location_name),
    direction = as.factor(direction),
    `Tax Status` = as.factor(`Tax Status`)  # Ensure backticks for column with space
  )
```

```{r}
filtered_data <- merged_data %>%
  filter(`Tax Status` == "Total")
```

```{r, fig.show='show', fig.cap="Interactive plot comparing hourly traffic volume patterns on weekdays versus weekends"}
# Create hour variable for plotting
filtered_data <- filtered_data %>%
  mutate(hour = as.numeric(format(time_start, "%H")))

# Prepare the data
filtered_data$weekday_label <- ifelse(filtered_data$is_weekend == 1, "Weekend", "Weekday")

hourly_summary <- filtered_data %>%
  group_by(weekday_label, hour) %>%
  summarise(avg_volume = mean(volume_15min, na.rm = TRUE), .groups = "drop")

# Interactive scatter plot
plot1 <- ggplot(hourly_summary, aes(x = hour, y = avg_volume, color = weekday_label)) +
  geom_point(size = 2) +
  geom_line() +
  labs(
    title = "Average Traffic Volume by Hour: Weekday vs Weekend",
    x = "Hour of Day",
    y = "Average 15-min Volume",
    color = "Day Type"
  ) +
  theme_minimal()

ggplotly(plot1)
```

```{r, fig.show='show', fig.cap="Interactive Leaflet map of average traffic volume by location in Toronto (2018–2022)"}
# Average volume by location
location_map_data <- filtered_data %>%
  group_by(location_name, latitude, longitude) %>%
  summarise(avg_volume = mean(volume_15min, na.rm = TRUE)) %>%
  ungroup()

# Create Leaflet map
leaflet(location_map_data) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude, lat = ~latitude,
    radius = ~sqrt(avg_volume), color = "blue", stroke = FALSE,
    fillOpacity = 0.6,
    popup = ~paste("Location:", location_name,
                   "<br>Avg Volume:", round(avg_volume, 1))
  ) %>%
  setView(lng = -79.38, lat = 43.65, zoom = 11)
```

```{r, fig.show='show', fig.cap="3D scatter plot showing the relationship between hour of day, fuel price, and 15-minute traffic volume (sample of 1,000 observations)"}
# Sample data from your actual dataset
plot_data <- filtered_data %>%
  sample_n(1000) %>%  # subset if too large
  mutate(hour = as.numeric(hour),
         fuel_price = Toronto,
         volume = volume_15min)

# 3D scatter
plot_ly(
  data = plot_data,
  x = ~hour, y = ~fuel_price, z = ~volume,
  type = "scatter3d", mode = "markers",
  marker = list(size = 3, color = ~volume, colorscale = "Viridis"),
  hoverinfo = "text",
  text = ~paste("Hour:", hour, "<br>Fuel Price:", fuel_price, "<br>Volume:", volume)
) %>%
  layout(title = "3D Scatter: Hour vs Fuel Price vs Volume")
```

## Results

### Modeling

This project fits two types of models:

**Statistical**: Generalized Additive Models (GAM)

**Machine Learning**: Regression Tree, Bagging, Random Forest, Boosting, XGBoost

### Best GAM Model

```{r}
filtered_data <- filtered_data %>%
  mutate(log_volume = log(volume_15min + 1))

filtered_data <- filtered_data %>%
  mutate(
    weekday = weekdays(actual_date),
    is_weekend = ifelse(weekday %in% c("Saturday", "Sunday"), 1, 0),
    hour = as.numeric(format(time_start, "%H")),
    pandemic = ifelse(actual_date >= as.Date("2020-03-01") & actual_date <= as.Date("2022-03-01"), 1, 0)
  )

```

```{r}
# Run this if you haven’t already fitted gam_model2
library(mgcv)
gam_model2 <- gam(log_volume ~ 
                    s(hour, bs = "cr") + 
                    s(Toronto, bs = "cr") +
                    is_weekend * pandemic,
                  data = filtered_data)
```

```{r}
# Summary of parametric terms
model2_summary <- summary(gam_model2)
kable(model2_summary$p.table, caption = "Parametric Terms (GAM Model 1)")
```

**Interpretation**: This model includes hour, fuel price, weekend indicator, pandemic period, and their interaction. Based on its EDF, GCV, and adjusted R², it offers the best trade-off between complexity and interpretability.

---

### Best Machine Learning Model: XGBoost

```{r}
# Train/test split
set.seed(48)
train_indices <- sample(1:nrow(filtered_data), 0.7 * nrow(filtered_data))
train_data <- filtered_data[train_indices, ]
test_data <- filtered_data[-train_indices, ]
```


```{r}
# If not already done
library(caret)
library(xgboost)
library(doParallel)

cl <- makeCluster(4)
registerDoParallel(cl)

train_control <- trainControl(method = "cv", number = 10)
tune_grid <- expand.grid(
  nrounds = seq(100, 500, 100),
  max_depth = c(2, 4, 6),
  eta = c(0.05, 0.1, 0.3),
  gamma = 0,
  colsample_bytree = 1,
  min_child_weight = 1,
  subsample = 1
)

xgb_model <- train(
  log_volume ~ hour + Toronto + is_weekend + pandemic,
  data = train_data,
  method = "xgbTree",
  trControl = train_control,
  tuneGrid = tune_grid,
  na.action = na.omit
)

stopCluster(cl)
```

```{r}
# Display best hyperparameters
kable(xgb_model$bestTune, caption = "Best Hyperparameters for XGBoost")
```

```{r}
# Variable importance plot
plot(varImp(xgb_model), main = "XGBoost Variable Importance")
```

**Interpretation**: XGBoost outperformed all other models on prediction accuracy, achieving the lowest test MSE.

---

### Model Comparison Table

```{r}
# --- Regression Tree (Pruned) ---
set.seed(123)
tree_model <- rpart(
  log_volume ~ hour + Toronto + is_weekend + pandemic,
  data = train_data,
  method = "anova",
  control = rpart.control(cp = 0)
)

cptable <- tree_model$cptable
desired_terminal_nodes <- 5
target_splits <- desired_terminal_nodes - 1
chosen_cp <- cptable[which(cptable[, "nsplit"] == target_splits), "CP"]
if(length(chosen_cp) == 0){
  chosen_cp <- cptable[which.min(abs(cptable[, "nsplit"] - target_splits)), "CP"]
}
pruned_tree <- prune(tree_model, cp = chosen_cp)

# --- Bagging ---
set.seed(123)
bagging_model <- randomForest(
  log_volume ~ hour + Toronto + is_weekend + pandemic,
  data = train_data,
  mtry = 4,
  ntree = 500,
  importance = TRUE
)

# --- Random Forest ---
set.seed(123)
rf_model <- randomForest(
  log_volume ~ hour + Toronto + is_weekend + pandemic,
  data = train_data,
  mtry = 2,
  ntree = 500,
  importance = TRUE
)

# --- Boosting ---
set.seed(301)
final_boost <- gbm(
  log_volume ~ hour + Toronto + is_weekend + pandemic,
  data = train_data,
  distribution = "gaussian",
  n.trees = 1000,
  interaction.depth = 3,
  shrinkage = 0.1,
  verbose = FALSE
)

# --- XGBoost ---
cl <- makeCluster(4)
registerDoParallel(cl)

train_control <- trainControl(method = "cv", number = 10, allowParallel = TRUE)

tune_grid <- expand.grid(
  nrounds = seq(100, 500, by = 100),
  max_depth = c(2, 4, 6),
  eta = c(0.05, 0.1, 0.3),
  gamma = 0,
  colsample_bytree = 1,
  min_child_weight = 1,
  subsample = 1
)

set.seed(301)
xgb_model <- train(
  log_volume ~ hour + Toronto + is_weekend + pandemic,
  data = train_data,
  method = "xgbTree",
  trControl = train_control,
  tuneGrid = tune_grid,
  na.action = na.omit,
  verbosity = 0
)

stopCluster(cl)
```

```{r}
# Ensure test_data and all models are defined
pred_tree <- predict(pruned_tree, newdata = test_data)
mse_tree <- mean((pred_tree - test_data$log_volume)^2)

pred_bagging <- predict(bagging_model, newdata = test_data)
mse_bagging <- mean((pred_bagging - test_data$log_volume)^2)

pred_rf <- predict(rf_model, newdata = test_data)
mse_rf <- mean((pred_rf - test_data$log_volume)^2)

pred_boost <- predict(final_boost, newdata = test_data, n.trees = 1000)
mse_boost <- mean((pred_boost - test_data$log_volume)^2)

pred_xgb <- predict(xgb_model, newdata = test_data)
mse_xgb <- mean((pred_xgb - test_data$log_volume)^2)

mse_results <- data.frame(
  Method = c("Regression Tree", "Bagging", "Random Forest", "Boosting", "XGBoost"),
  Test_MSE = round(c(mse_tree, mse_bagging, mse_rf, mse_boost, mse_xgb), 4)
)

kable(mse_results, caption = "Test MSE for Each Regression Model")
```

